Multilevel selection in causal models: the non-ergodicity of evolutionary and probabilistic theories as the general driver of the emergence of cooperation and specialization.

### 2022-01-21 article.tex
Revision of introduction and methodology sections

\en{The reason why an advantage in favor of cooperation and specialization arises in our simple causal model is due to the multiplicative (non-ergodic) nature of probabilistic theory and its isomorphism with evolutionary theory. }%
\es{El motivo por el cual surge una ventaja a favor de la cooperación y la especialización en simple modelo causal se debe a la naturaleza multiplicativa (no-ergódica) de la teoría de la probabililidad y a su isomorfismo con la teoría de la evolución. }%
%
% \en{Even if our causal model does not assume any kind of process to update resources, probabilistic inference is proportional to the resources obtained through a multiplicative process. }%
% \es{Incluso si nuestro modelo causal no presupone ningún tipo de proceso para actualizar los recursos, la inferencia probabilistica es proporcional a los recursos obtenidos a través de un proceso multiplicativo. }%
% %
\en{The multiplicative updating of the probabilities of individuals, which arises naturally from applying the rules of probability to the causal model, is in line with the long-established idea in evolutionary theory that the growth of lineages follow multiplicative processes. }%
\es{La actualización multiplicativa de las probabilidades de los individuos, que surge naturalmente de aplicar las reglas de la probabilidad al modelo causal, está en línea con la idea largamente establecida en la teoría de la evolución de que el crecimiento de los linajes siguen procesos multiplicativos. }% 
%

@@

\es{Sin embargo, las condiciones para que esto se cumpla son muy restrictivas y no se satisfacen para el caso de los procesos multiplicativos. }%
\en{To calculate the characteristic growth rate $g$, we first express the product as follows, }%
\es{Para calcular la tasa de crecimiento caracterísitica $g$, primero expresaramos la productoria de la siguiente manera, }%
\omega(T) = \prod^T_{t=1} f(a(t)) = f(\text{\en{head}\es{cara}})^{n_1} f(\text{\en{tail}\es{sello}})^{n_2}

@@

\en{In the limit, $T \rightarrow \infty$ all individual trajectories will be determined by the same characteristic growth rate $g$. }%
\es{En el límite, $T \rightarrow \infty$ todas las trayectorias individuales estarán determinadas por la misma tasa de crecimiento caracterísitica $g$. }%

@@

\en{Where the frequencies $\frac{n_1}{T}$ and $\frac{n_2}{T}$ in the limit $T \rightarrow \infty$ are equal to the probabilities of occurrence of the system states. }%
\es{Donde las frecuencias $\frac{n_1}{T}$ y $\frac{n_2}{T}$ en el límite $T \rightarrow \infty$ son iguales a las probabilidades de ocurrencia de los estados del sistema. }%

@@

\en{An important property of the geometric mean is that its value is always less than the expected value (or arithmetic mean). }%
\es{Una propiedad importante de la media geométrica es que su valor siempre es menor al valor esperado (o media aritmética). }%

@@

\en{As a consequence of the non-ergodicity of multiplicative processes, fluctuations have a negative effect on individual growth rates, which can be reduced through mutual cooperation. }%
\es{Como consecuencia de la no-ergodicidad de los procesos multiplicativos, las fluctuaciones tienen un efecto negativo en las tasas de crecimiento individuales, las cuales pueden ser reducidas a través de la mutua cooperación. }%
%
%Parafraseando a Den Boer~\cite{denBoer1968-spreadingRisk}, la  supervivencia de una población depende de la distribución del riesgo dentro de la población y entre las poblaciones de diferentes especies.
\en{Yaari-Solomon and Peters-Adamou~\cite{yaari2010-cooperationEvolution, peters-cooperation2019.03.04} (hereinafter Yaari and Peters) considers the following cooperative strategy and analyzes the consequences it has on the growth rate of the agents. }%
\es{Yaari-Solomon and Peters-Adamou~\cite{peters-cooperation2019.03.04} (de aquí en adelante Yaari y Peters) consideran la siguiente estrategia cooperativa y analiza la consecuencias que tiene sobre la tasa de crecimiento de los agentes. }%
%
\begin{figure}[H]
\centering
\scalebox{0.75}{
\tikz{

@@

}
}
\caption{
\en{Agents start with the same initial resources. They then grow independently according to the equation \ref{eq:estrategia_base}. They then cooperate by pooling and sharing their resources. }%
\es{Los agentes comienzan con los mismos recursos iniciales. Luego crecen independientemente de acuerdo con la ecuaci\'on \ref{eq:estrategia_base}. Luego cooperan poniendo sus recursos en un fondo común que dividen en partes iguales. }%
}
\label{fig:protocolo}
\end{figure}
%
\en{Fully cooperative groups reduce their fluctuations, generating an increase in the growth rate of all their members. }%
\es{Grupos enteramente cooperadoras reducen sus fluctuaciones, generando un aumento en la tasa de crecimiento de todos sus miembros. }%
%
\en{In Figure \ref{fig:ergodicity_cooperation} we show the trajectory of an agent in a cooperating group of size 33. }%
\es{En la figura \ref{fig:ergodicity_cooperation} mostramos la trayectoria de un agente en un grupo cooperador de tamaño 33. }%
%
\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/pdf/ergodicity_cooperation.pdf}
    \end{subfigure}
    \caption{
    \en{The resources of an individual from a group of 33 agents who share their wealth after each iteration (green line), approaches the expected value (black line). As a visual reference, we show the growth rate of the individuals (blue line). }%
    \es{Los recursos de un individuo de una grupo de 33 agentes que comparten su riqueza luego de cada iteración (recta verde), se pega al valor esperado (recta negra).
    Como referencia visual, dejamos la tasa de crecimiento de los individuos (recta azul). }%
    }
    \label{fig:ergodicity_cooperation}
\end{figure}
%
% \paragraph{Conclusión Ole Peters} (La ventaja de la cooperación)\textbf{.}
\en{Individuals in fully cooperative groups achieve growth rates equivalent to the average of system states, which in non-ergodic systems is always higher than individual growth rates. }%
\es{Los individuos de los grupos enteramente cooperadoras logran acceder a tasas de crecimiento equivalentes al promedio de estados del sistema, que en los sistemas no-ergódicos es siempre superior que la tasas de crecimiento individual. }%
%
\en{Ole Peters believs that the increase in the growth rate is sufficient argument to demonstrate the evolutionary advantage of cooperation, which he proposes as the main explanation for evolutionary transitions. }%
\es{Ole Peters considera que el aumento de la tasa de crecimiento es argumento suficiente para demostrar la ventaja evolutiva de la cooperación, lo que propone como principal explicación de las transiciones evolutivas. }%
%
\en{However, he does not consider the problem of defection, who says ``our cooperators are unable to break the cooperative pact''. }%
\es{Sin embargo no considera el problema de la deserción, quien dice ``our cooperators are unable to break the cooperative pact''. }%
%
\en{This does not seem to be a minor problem, considering the temptation to stop contributing to the common fund while continuing to receive its benefits. }%
\es{No parece ser un problema menor, teniendo en cuenta la tentación de dejar de aportar al fondo común mientras se siguen recibiendo sus beneficios. }%

@@

\en{To demonstrate the evolutionary advantage of cooperation in the presence of defection it is necessary to consider selection at both the individual and group level. }%
\es{Para demostrar la ventaja evolutiva de la cooperción en presencia de desertores es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{Yaari and Solomon~\cite{yaari2010-cooperationEvolution} already suggest a verbal argument for group selection: while there would be a temptation to defect, fully cooperative groups will do better. }%
\es{Yaari y Solomon~\cite{yaari2010-cooperationEvolution} sugieren ya un argumento verbal de la selección de grupo: si bien habría una tentación para desertar, a los grupos enteramente cooperadores le irá mejor. }%
%
\en{However, they do not provide this argument with a formal demonstration. }%
\es{Sin embargo no acompañan este argumento con una demostración formal. }%

\en{The co-author of the concept of evolutionary transitions (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models~\cite{czegel2019-bayesianEvolution}, making use of the isomorphism between evolutionary theory and Bayesian inference~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
\es{El co-autor del concepto de transiciones evolutivas (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos~\cite{czegel2019-bayesianEvolution}, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
%
\en{However, this work does not provide any model that performs multilevel selection, so the proposal remains open. }%
\es{Sin embargo, este trabajo no proveé ningun modelo que realice selección multinivel, por lo que la propuesta sigue abierta. }%

\en{In this paper we demonstrate the evolutionary advantage of cooperation in the presence of defection using a Bayesian hierarchical model representing evolution under multilevel selection. }%
\es{En este trabajo demostramos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo jerárquico bayesiano que represente evolución bajo selección multinivel. }%
%
\en{To the best of our knowledge, our work would be the first to develop a Bayesian hierarchical model to solve an evolution problem under multilevel selection. }%
\es{Hasta donde sabemos, nuestro trabajo sería el primero en desarrollar un modelo jerárquico bayesiano para resolver un problema de evolución bajo selección multinivel. }%
%
% \en{With this model we demonstrate, as Ole Peters intended, that the advantage of cooperation (now in the presence of defection) is a consequence of the most basic and fundamental assumption of evolutionary theory: that evolutionary processes are multiplicative and noisy. }%
% \es{Con este modelo demostramos, como pretendía Ole Peters, de que la ventaja de la cooperación (ahora en presencia de deserción) es consecuencia del supuesto más básico y fundamental de la teoría de la evolución: que los procesos evolutivos son multiplicativos y ruidosos. }%

\paragraph{Especialización}

% \en{According to Ole Peters, an important evolutionary phenomenon on which his analysis can shed new light is the transition to multicellularity. }%
% \es{Según Ole Peters, un importante fenómeno evolutivo sobre el que su análisis puede arrojar nueva luz es la transición a la multicelularidad. }%
%
\en{The evolutionary transition from unicellularity to multicellularity does not only involve cooperation, it also involves a specialization of cells: from protists to plants, fungi and animals. }%
\es{La transición evolutiva de la unicelularidad a la multiclelularidad no involucra sólo cooperación, involucra también una especialización de las células: de los los protistas a las plantas, hongos y animales. }%
%
\en{Then, to explain evolutionary transitions, it is necessary to demonstrate the evolutionary advantage of cooperation in the presence of defection, but also the advantage of specialization. }%
\es{Luego, para explicar las transiciones evolutivas es necesario demostrar la ventaja evolutiva de la cooperación en presencia de deserción, pero también la ventaja de la especiliazación. }%

% Parrafo
% 
% \en{However, Ole Peters believes that the cause of the cooperation advantage (the assumption of non-ergodicity of noisy multiplicative processes) does not allow explaining the specialization advantage. }%
% \es{Sin embargo, Ole Peters cree que la causa de la ventaja de la cooperación (el supuesto de la no-ergodicidad de los procesos multiplicativos ruidosos) no permite explicar la ventaja de la especialización. }%
% %
% \en{Ole Peters believes that specialization is a too complex property to produce a benefit in simple aggregations of, for example, two agents. }
% \es{Ole Peters considera a la especialización como una caracterísitica demasiado compleja para que produzca un beneficio en agregaciones simples de por ejemplo dos agentes. }%
% %
% \en{He says the ``[advantage of cooperation] it may explain the transition from single cells to bicellular organisms, too small and simple to benefit from new function or specialization.'' }
% \es{Dice que la ventaja de la cooperación: ``it may explain the transition from single cells to bicellular organisms, too small and simple to benefit from new function or specialization.'' }%
% %
% \en{He believes that other explanations of evolutionary transitions, such as the emergence of specialization, are implausible. }%
% \es{El cree que otras explicaciones de las transiciones evolutivas, como la emergencia de la especialización, son implausibles. }%

\en{In this paper we show that the same minimal assumptions used by Yaari and Peters to explain the evolutionary advantage of cooperation also allow us to explain the evolutionary advantage of specialization, even for the simplest aggregations of two agents. }%
\es{En este trabajo demostramos que los mismos supuestos mínimos que utilizam Yaari y Peters para explicar la ventaja evolutiva de la cooperación, permiten también explicar la ventaja evolutiva de especialización, incluso para las agregaciones más simples de dos agentes. }%
%
\en{We show that, once cooperation emerges, individually maladapted strategies (specialists) manage to outperform, even organized in small groups, both individually well-adapted strategies (generalists) and their cooperative groups of infinite size. }%
\es{Mostramos que, una vez que la cooperación emerge, las estrategias individualmente mal adaptadas (especialistas) consigen superar, incluso organizadas en grupos pequeños, tanto a las estrategias bien adaptas individualmemte (generalistas) como a sus grupos cooperativos de tamaño infinito. }%

\en{It is extraordinary that such a simple assumption has such fundamental conclusions for understanding the complexity of life. }%
\es{Es extraordinario que un supuesto tan simple tenga conclusiones tan fudamentales para entender la complejidad de la vida. }%
%
% \en{Because the evolutionary advantage of cooperation and specialization is a direct consequence of the non-ergodicity of the multiplicative processes to which life is subject, we support Ole Peters' idea of considering it the first cause of major evolutionary transitions. }%
% \es{Debido a que la ventaja evolutiva de la cooperación y la especialización es consecuencia directa de la no-ergodicidad de los procesos multiplicativos a los que está sujeto la vida, apoyamos la idea de Ole Peters de consideradala la primera causa de las transiciones evolutivas mayores. }%

\section{Metodología}

\en{In this section we present the ismorphism between probability theory and evolutionary theory. }%
\es{En esta sección presentamos el ismorfismo entre la teoría de la probabilidad y la teoría de la evolución. }%
%
\en{In the Results section we will demonstrate the evolutionary advantage of cooperation and specialization using a Bayesian hierarchical model representing evolution under multilevel selection. }%
\es{En la sección Reseultados demostraremos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo jerárquico bayesiano que represente evolución bajo selección multinivel. }%
%

\subsection{Teoría de la probabilidad y los modelos causales}

\en{Probability theory is currently the most widely used approach to computing uncertainty. }%
\es{La teoría de la probabilidad es el enfoque más utilizado en la actualidad para computar la incertidumbre. }%
%
\en{Its rules have been formally derived from conceptually distinct and mutually independent axiomatic systems~\cite{halpern2017-RAU2}, which is one of the strong points in its favor. }%
\es{Sus reglas han sido derivadas formalmente a partir de sistemas axiomáticos conceptualmente distintos e independientes entre sí~\cite{halpern2017-RAU2}, lo cual es uno de los punto fuertes a su favor. }%
%
\en{But perhaps more importantly, its strict application maximizes uncertainty given empirical and formal information (data and causal models)~\cite{jaynes2003-bookProbabilityTheory}, the source of validation for empirical sciences. }%
\es{Pero quizás más importante sea que su aplicación estricta maximiza la incertidumbre dada la información empírica y formal (datos y modelos causales)~\cite{jaynes2003-bookProbabilityTheory}, fuente de validación de las ciencias empíricas. }%

\en{The whole theory of probability can be summarized in two rules: the \ref{eq:sum_rule} and the ~\ref{eq:product_rule}. }%
\es{Toda la teoría de la probabilidad puede resumirse en dos reglas: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}. }%
%
\en{The \ref{eq:sum_rule} states that any marginal distribution can be obtained by integrating or summing the joint distribution. }%
\es{La \ref{eq:sum_rule} afirma que cualquier distribución marginal se puede obtener integrando o sumando la distribución conjunta. }%
%
\begin{equation} \label{eq:sum_rule}
 \tag{\en{sum rule}\es{regla de la suma}}
 P(x) = \sum_{y} P(x,y) \ \ \ \ \ \text{or} \ \ \ \ \ p(x) = \int p(x,y) \, dy
\end{equation}
%
\en{Where $P(\cdot)$ and $p(\cdot)$ represent discrete and continuous probability distributions respectively. }%
\es{Donde $P(\cdot)$ y $p(\cdot)$ representan distribuciones de probabilidad discretas y continuas respectivamente. }%
%
\en{On the other hand, the \ref{eq:product_rule} states that any joint distribution can be expressed as the product of one-dimensional conditional distributions. }%
\es{Por su parte, la \ref{eq:product_rule} se\~nala que cualquier distribuci\'on conjunta puede ser expresada como el producto de distribuciones condicionales uni-dimensionles. }%
%
\begin{equation}\label{eq:product_rule}
\tag{\en{product rule}\es{regla del producto}}
 P(x,y,z) = P(x|y,z) P(y|z) P(z)
\end{equation}
%
\en{This expression can be simplified when we eliminate from the conditional the variables that do not have any effect on the function. }%
\es{Esta expresión se puede simplificar cuando eliminamos del condicional las variables que no afectan a la función. }%
%
\en{The minimum set of variables that appear on the right side of the conditional are said to be the ``parents'' of the variable that appears on the left side of the conditional. }%
\es{Al conjunto mínimo de variables que aparecen del lado derecho del condicional decimos que son los ``padres'' de la variable que aparece del lado izquierdo del condicional. }%

>>>>

\en{The reason why an advantage in favor of cooperation and specialization arises in our simple causal model is due to the multiplicative (non-ergodic) nature of probability theory and its isomorphism with evolutionary theory. }%
\es{El motivo por el cual surge una ventaja a favor de la cooperación y la especialización en simple modelo causal se debe a la naturaleza multiplicativa (no-ergódica) de la teoría de la probabililidad y a su isomorfismo con la teoría de la evolución. }%
%
% \en{Even if our causal model does not assume any kind of process to update resources, probabilistic inference is proportional to the resources obtained through a multiplicative process. }%
% \es{Incluso si nuestro modelo causal no presupone ningún tipo de proceso para actualizar los recursos, la inferencia probabilistica es proporcional a los recursos obtenidos a través de un proceso multiplicativo. }%
% %
\en{The multiplicative updating of the probabilities of individuals, which arises naturally from applying the rules of probability to the causal model, is in line with the long-established idea in evolutionary theory that the growth of lineages follow multiplicative processes. }%
\es{La actualización multiplicativa de las probabilidades de los individuos, que surge naturalmente de aplicar las reglas de la probabilidad al modelo causal, está en línea con la idea largamente establecida en la teoría de la evolución de que el crecimiento de los linajes siguen procesos multiplicativos. }% 
%

@@

\es{Sin embargo, las condiciones para que esto se cumpla son muy restrictivas y no se satisfacen en el caso de los procesos multiplicativos. }%
\en{To compute the growth rate $g$, we first express the product as follows, }%
\es{Para calcular la tasa de crecimiento $g$, primero expresaramos la productoria de la siguiente manera, }%
\omega(T) = \prod^T_{t=1} f(\Aa(t)) = f(\text{\en{head}\es{cara}})^{n_1} f(\text{\en{tail}\es{sello}})^{n_2}

@@

\en{In the limit, $T \rightarrow \infty$ all individual trajectories will be determined by the same growth rate $g$. }%
\es{En el límite, $T \rightarrow \infty$ todas las trayectorias individuales estarán determinadas por la misma tasa de crecimiento $g$. }%

@@

\en{Where the frequencies $\frac{n_1}{T}$ and $\frac{n_2}{T}$ in the limit $T \rightarrow \infty$ are equal to the probabilities of the environmental states $p$. }%
\es{Donde las frecuencias $\frac{n_1}{T}$ y $\frac{n_2}{T}$ en el límite $T \rightarrow \infty$ son iguales a las probabilidades de los estados ambientales $p$. }%

@@

\en{The geometric mean is always less than the arithmetic mean (or expected value). }%
\es{La media geométrica es siempre es menor a la media aritmética (o valor esperado). }%

@@

\en{As a consequence of the non-ergodicity of multiplicative processes, fluctuations have a negative effect on individual growth rates. }%
\es{Como consecuencia de la no-ergodicidad de los procesos multiplicativos, las fluctuaciones tienen un efecto negativo en las tasas de crecimiento individuales. }%
%
%Parafraseando a Den Boer~\cite{denBoer1968-spreadingRisk}, la  supervivencia de una población depende de la distribución del riesgo dentro de la población y entre las poblaciones de diferentes especies.
\en{Yaari-Solomon~\cite{yaari2010-cooperationEvolution} and Peters-Adamou~\cite{peters-cooperation2019.03.04} (hereinafter Yaari and Peters) study the consequences that the following cooperative strategy has on the agents' growth rate. }%
\es{Yaari-Solomon~\cite{yaari2010-cooperationEvolution} and Peters-Adamou~\cite{peters-cooperation2019.03.04} (de aquí en adelante Yaari y Peters) estudian las consecuencias que la siguiente estrategia cooperativa tiene sobre la tasa de crecimiento de los agentes. }%
%
\begin{figure}[H]
\centering
\scalebox{0.75}{
\tikz{

@@

}
}
\caption{
\en{The agents start with the same initial resources, update them independently according to the equation \ref{eq:estrategia_base}, and finally redistribute it in equal parts. }%
\es{Los agentes comienzan con los mismos recursos iniciales, los actualizan independientemente de acuerdo con la ecuación \ref{eq:estrategia_base}, y finalmente lo redistribuyen en partes iguales. }%
}
\label{fig:protocolo}
\end{figure}
%
\en{In Figure \ref{fig:ergodicity_cooperation} we show the trajectory of an agent in a cooperating group of size 33. }%
\es{En la figura \ref{fig:ergodicity_cooperation} mostramos la trayectoria de un agente en un grupo cooperador de tamaño 33. }%
%
\begin{figure}[ht!]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/pdf/ergodicity_cooperation.pdf}
    \end{subfigure}
    \caption{
    \en{The resources of an individual in a cooperative group of size 33 (green line) approaches the arithmetic mean (the black line below the green line).
    As a visual reference we show the geometric mean that drives the growth rate of individuals (blue line). }%
    \es{Los recursos de un individuo en un grupo cooperativo de tamaño 33 (recta verde) se aproxima a la media aritmética (la recta negra debajo de la recta verde).
    Como referencia visual mostramos la media geométrica que gobierna la tasa de crecimiento de los individuos (recta azul). }%
    }
    \label{fig:ergodicity_cooperation}
\end{figure}
%
\en{Fully cooperative groups reduce their fluctuations, generating an increase in the growth rate of all their members. }%
\es{Grupos enteramente cooperadoras reducen sus fluctuaciones, generando un aumento en la tasa de crecimiento de todos sus miembros. }%
% 
% \en{Individuals in fully cooperative groups achieve growth rates equivalent to the average of system states, which in non-ergodic systems is always higher than individual growth rates. }%
% \es{Los individuos de los grupos enteramente cooperadoras logran acceder a tasas de crecimiento equivalentes al promedio de estados del sistema, que en los sistemas no-ergódicos es siempre superior que la tasas de crecimiento individual. }%
%
\en{Peters-Amadou~\cite{peters-cooperation2019.03.04} believe that such increase in the growth rate is sufficient argument to demonstrate the evolutionary advantage of cooperation. }%
\es{Peters-Amadou~\cite{peters-cooperation2019.03.04} consideran que el aumento de la tasa de crecimiento es argumento suficiente para demostrar la ventaja evolutiva de la cooperación. }%
%
\en{However, they do not consider the problem of defection, who say ``our cooperators are unable to break the cooperative pact''. }%
\es{Sin embargo ellos no consideran el problema de la deserción, quienes dicen ``our cooperators are unable to break the cooperative pact''. }%
%
\en{This does not seem to be a minor problem, considering the temptation to stop contributing to the common fund while continuing to receive its benefits. }%
\es{No parece ser un problema menor, teniendo en cuenta la tentación de dejar de aportar al fondo común mientras se siguen recibiendo sus beneficios. }%

@@

\en{To demonstrate the evolutionary advantage of cooperation in the presence of defection it is necessary to consider selection at both the individual and group level. }%
\es{Para demostrar la ventaja evolutiva de la cooperción en presencia de desertores es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{Yaari and Solomon~\cite{yaari2010-cooperationEvolution} already suggest a verbal argument for group selection: while there would be a temptation to defect, fully cooperative groups will do better. }%
\es{Yaari y Solomon~\cite{yaari2010-cooperationEvolution} sugieren ya un argumento verbal de la selección de grupo: si bien habría una tentación para desertar, a los grupos enteramente cooperadores le irá mejor. }%
%
\en{However, they do not provide this argument with a formal proof. }%
\es{Sin embargo no acompañan este argumento con una demostración formal. }%
% Parrafo
\en{The co-author of the concept of evolutionary transitions (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models~\cite{czegel2019-bayesianEvolution}, making use of the isomorphism between evolutionary theory and Bayesian inference~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
\es{El co-autor del concepto de transiciones evolutivas (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos~\cite{czegel2019-bayesianEvolution}, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
%
\en{However, this work does not provide any model that performs multilevel selection, so the proposal remains open. }%
\es{Sin embargo, este trabajo no proveé ningun modelo que realice selección multinivel, por lo que la propuesta sigue abierta. }%
% Parrafo
\en{In this paper we demonstrate the evolutionary advantage of cooperation in the presence of defection using a Bayesian hierarchical model representing evolution under multilevel selection. }%
\es{En este trabajo demostramos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo jerárquico bayesiano que represente evolución bajo selección multinivel. }%
%
\en{To the best of our knowledge, our work would be the first to develop a Bayesian hierarchical model to solve an evolution problem under multilevel selection. }%
\es{Hasta donde sabemos, nuestro trabajo sería el primero en desarrollar un modelo jerárquico bayesiano para resolver un problema de evolución bajo selección multinivel. }%
%
% \en{With this model we demonstrate, as Ole Peters intended, that the advantage of cooperation (now in the presence of defection) is a consequence of the most basic and fundamental assumption of evolutionary theory: that evolutionary processes are multiplicative and noisy. }%
% \es{Con este modelo demostramos, como pretendía Ole Peters, de que la ventaja de la cooperación (ahora en presencia de deserción) es consecuencia del supuesto más básico y fundamental de la teoría de la evolución: que los procesos evolutivos son multiplicativos y ruidosos. }%

\paragraph{\en{Specialization}\es{Especialización}}

\en{Peters-Amadou propose multiplicative processes as the main explanation of major evolutionary transitions. }%
\es{Peters-Amadou proponen a los procesos multiplicativos como la principal explicación de la transiciones evolutvas mayores. }%
\en{However, they believe that multiplicative processes can only explain the advantage of cooperation, but not the advantage of specialization. }%
\es{Sin embargo, ellos creen que los procesos multiplicativos sólo pueden explicar la ventaja de la cooperación, pero no la ventaja de la especialización. }%
% \en{Ole Peters believes that specialization is a too complex property to produce a benefit in simple aggregations of, for example, two agents. }
% \es{Ole Peters considera a la especialización como una caracterísitica demasiado compleja para que produzca un beneficio en agregaciones simples de por ejemplo dos agentes. }%
% \en{He says the ``[advantage of cooperation] it may explain the transition from single cells to bicellular organisms, too small and simple to benefit from new function or specialization.'' }
% \es{Dice que la ventaja de la cooperación: ``it may explain the transition from single cells to bicellular organisms, too small and simple to benefit from new function or specialization.'' }%
% \en{He believes that other explanations of evolutionary transitions, such as the emergence of specialization, are implausible. }%
% \es{El cree que otras explicaciones de las transiciones evolutivas, como la emergencia de la especialización, son implausibles. }%
% \en{According to Ole Peters, an important evolutionary phenomenon on which his analysis can shed new light is the transition to multicellularity. }%
% \es{Según Ole Peters, un importante fenómeno evolutivo sobre el que su análisis puede arrojar nueva luz es la transición a la multicelularidad. }%
% \en{But, to explain evolutionary transitions, it is necessary to demonstrate the evolutionary advantage of cooperation in the presence of defection, but also the advantage of specialization. }%
% \es{Luego, para explicar las transiciones evolutivas es necesario demostrar la ventaja evolutiva de la cooperación en presencia de deserción, pero también la ventaja de la especiliazación. }%
% \en{However evolutionary transitions not only involve cooperation, it also involves specialization. }%
% \es{Sin embargo, las transiciones evolutivas no involucran sólo cooperación, involucra también especialización. }%
\en{In this paper we show that the same minimal assumptions used by Yaari and Peters to explain the evolutionary advantage of cooperation also allow us to explain the evolutionary advantage of specialization, even for the simplest aggregations of two agents. }%
\es{En este trabajo demostramos que los mismos supuestos mínimos que utilizam Yaari y Peters para explicar la ventaja evolutiva de la cooperación, permiten también explicar la ventaja evolutiva de especialización, incluso para las agregaciones más simples de dos agentes. }%
\en{Once cooperation emerges, individually maladapted strategies (specialists) manage to outperform, even organized in small groups, both individually well-adapted strategies (generalists) and their cooperative groups of infinite size. }%
\es{Una vez que la cooperación emerge, las estrategias individualmente mal adaptadas (especialistas) consigen superar, incluso organizadas en grupos pequeños, tanto a las estrategias bien adaptas individualmemte (generalistas) como a sus grupos cooperativos de tamaño infinito. }%
\en{It is extraordinary that such an essential assumption, i.e. the multiplicative nature of probability and evolutionary theories, has such fundamental conclusions for understanding the emergence of the complexity of life. }%
\es{Es extraordinario que un supuesto tan esencial, i.e. la naturaleza multiplicativa de las teorías probabilisticas y evolutivas, tenga conclusiones tan fudamentales para entender la emergencia de la complejidad de la vida. }%
% \en{Because the evolutionary advantage of cooperation and specialization is a direct consequence of the non-ergodicity of the multiplicative processes to which life is subject, we support Ole Peters' idea of considering it the first cause of major evolutionary transitions. }%
% \es{Debido a que la ventaja evolutiva de la cooperación y la especialización es consecuencia directa de la no-ergodicidad de los procesos multiplicativos a los que está sujeto la vida, apoyamos la idea de Ole Peters de consideradala la primera causa de las transiciones evolutivas mayores. }%

\section{\en{Methodology}\es{Metodología}}\label{sec:methodology}

\en{In this section we present the isomorphism between probabilitic and evolutionary theory. }%
\es{En esta sección presentamos el isomorfismo entre las teorías de la probabilidad y de la evolución. }%
%
\en{In the \ref{sec:resuts} section we will poof the evolutionary advantage of cooperation and specialization using a Bayesian hierarchical model representing evolution under multilevel selection. }%
\es{En la sección \ref{sec:results} demostraremos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo jerárquico bayesiano que representa evolución bajo selección multinivel. }%
%

\subsection{\en{Probability theory and causal models}\es{Teoría de la probabilidad y modelos causales}}

\en{Probability theory is currently the most widely used approach for computing uncertainty. }%
\es{La teoría de la probabilidad es el enfoque más utilizado en la actualidad para computar la incertidumbre. }%
%
\en{Its rules have been derived from several axiomatic systems, conceptually distinct and independent of each other~\cite{halpern2017-RAU2}, which is one of the strong points in its favor. }%
\es{Sus reglas han sido derivadas a partir de varios sistemas axiomáticos, conceptualmente distintos e independientes entre sí,~\cite{halpern2017-RAU2}, lo cual es uno de los punto fuertes a su favor. }%
%
\en{But perhaps more importantly, its strict application ensures maximization of uncertainty given empirical and formal information (data and causal models)~\cite{jaynes2003-bookProbabilityTheory}, source of validation for empirical propositions. }%
\es{Pero quizás más importante es que su aplicación estricta garatiza la maximización de la incertidumbre dada la información empírica y formal (datos y modelos causales)~\cite{jaynes2003-bookProbabilityTheory}, fuente de validación para las proposiciones empíricas. }%

\en{Probability theory can be summarized in two rules: the \ref{eq:sum_rule} and the~\ref{eq:product_rule}. }%
\es{La teoría de la probabilidad puede resumirse en dos reglas: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}. }%
%
\en{The \ref{eq:sum_rule} states that any marginal distribution can be obtained by integrating or summing the joint distribution. }%
\es{La \ref{eq:sum_rule} afirma que cualquier distribución marginal se puede obtener integrando o sumando la distribución conjunta. }%
%
\begin{equation} \label{eq:sum_rule}
 \tag{\en{sum rule}\es{regla de la suma}}
 P(x) = \sum_{y} P(x,y) \ \ \ \ \ \text{or} \ \ \ \ \ p(x) = \int p(x,y) \, dy
\end{equation}
%
\en{Where usually $P(\cdot)$ and $p(\cdot)$ represent discrete and continuous probability distributions respectively. }%
\es{Donde usualmente $P(\cdot)$ y $p(\cdot)$ representan distribuciones de probabilidad discretas y continuas respectivamente. }%
%
\en{On the other hand, the \ref{eq:product_rule} states that any joint distribution can be expressed as the product of one-dimensional conditional distributions. }%
\es{Por su parte, la \ref{eq:product_rule} se\~nala que cualquier distribuci\'on conjunta puede ser expresada como el producto de distribuciones condicionales uni-dimensionles. }%
%
\begin{equation}\label{eq:product_rule}
\tag{\en{product rule}\es{regla del producto}}
 P(x,y,z) = P(x|y,z) P(y|z) P(z)
\end{equation}
%
\en{This expression can be simplified by eliminating from the conditionals the variables that do not affect the function. }%
\es{Esta expresión se puede simplificar eliminado de los condicionales las variables que no afectan a la función. }%
%
\en{The minimum set of variables that appear on the right side of the conditional are said to be the ``parents'' of the left side variable. }%
\es{Al conjunto mínimo de variables que aparecen del lado derecho del condicional decimos que son los ``padres'' de la variable de la izquierda. }%
